### 2011-06-14 thesis/semantic-mapping.tex
Fixes here and there.
Initial fixes of semantic mapping chapter.
"Spatial Knowledge Representation" is now clear one concept.
And dora specific implementation another.

There has been several efforts in the areas of artificial intelligence and robotics
in creating robots that are able to interact with humans and their environments.
One of the important aspects is to endow those robots with a deeper understanding
of human environments, not just for the purpose of navigation and obstacle avoidance, 
but also in terms of human semantics and functionality.

In that sense the need for semantic mapping arises, and it can be described as the
relation between spatial areas and a set of existent concepts used by humans to
describe them.
Humans associate concepts to areas that relate to the functionality of those. That
association is either useful by explaining where to find certain objects as well to
understand the types of activities to perform on it.
By being able to reason over those semantics an agent can be expected to better
understand human environments and increase its efficiency when performing tasks on them.

@@

This thesis motivation lies on developing methods for detecting knowledge gaps
on the semantic knowledge of a mobile agent. For that the semantic mapping system
proposed by \cite{pronobis2011phd} was used as a base.

\section{Dora Architecture Overview}
Pronobis presents a system architecture working on indoor environments using
non\hyp{}omnidirectional lasers and visual sensors, that by integrating cues such as geometry,
object presence and appearances is able to perform inference across any semantic property,
for example to handle place categorization.
The introduced semantic representation has not only been tested on real scenarios
and performance measured across several conditions~\cite{pronobis2007iros} but also shown
to be tailored to effectively solve tasks arising on mobile robotics~\cite{hanheide2011ijcai}.
Besides that the system architecture has been drawn with an high-focus on probabilistic
and uncertain reasoning, human interaction and life-long learning capabilities.

For that it was considered an excellent base defining not only the concepts and
requirements of an high-level semantic representation but also the information flow
between all the layers involved on the agent.
The system is implemented on \Gls{Dora} and a brief overview of its architecture is
given in the following sections.

\subsection{System organization}
The \Gls{Dora} system~\cite{hanheide2011ijcai} consists of several co-operating
sub-systems, all of which actively use or maintain the spatial knowledge representation
(see \autoref{fig:dora-architecture}):

\begin{figure}[h]
\centering
\includegraphics[width=0.7\textwidth]{figures/dora-architecture.pdf}
\caption{\label{fig:dora-architecture}Interaction of the sub-systems
         in Dora with special focus on the conceptual layer.}
\end{figure}

\subsubsection*{Sensory Layer}
The sensory layer is responsible for all the low-level sensors.
It maintains an immediate and exact description of the environment by using
data such as images from cameras, 2D\hyp{}laser scans and odometry from encoders in the
movement motors. The information provided by it is used on the higher layers to extract
tractable information.

\subsubsection*{Place Layer}
The place layer is responsible for discretizing the continuous low-level space into
a finite number of places by spreading several virtual places markers around the whole
environment and maintaining connectivity information between them.
This virtual places also handle information associating the robot view point. Together
with extracted properties they allow the system to perform a clever integration over the
room properties, avoiding sampling bias, for example coming due to the robot spend too
much time in a part of the room~\cite{pronobis2010ijrr}.

\subsubsection*{Categorical Layer}
The categorical layer is responsible for extracting high-level information from the
sensors. It has classifiers trained to detect several properties from the visual and laser
features captured on the sensory layer.
It is responsible for detecting objects, doors, categorize the geometry and the appearance.
The used methods and properties extracted by this layer are described in more detail in
\autoref{sec:dora-features}.

\subsubsection*{Conceptual Layer}
The conceptual layer merges the sensed properties and places structure together
with general knowledge under a unified probabilistic semantic representation: the
\emph{conceptual map}.
This conceptual map is the structure for handling the high-level semantic information
and performing inferences. It is the layer where the novelty detection methods described
on this thesis are meant to be integrated and is described in more detail in
\autoref{sec:conceptual-map}.

\subsection{Features}
\label{sec:dora-features}

\section{Conceptual Map}
\label{sec:conceptual-map}
As \gls{Dora} moves through the environment its \emph{conceptual layer} builds a structural and
probabilistic representation of the space instantiated as a \emph{graphical model}.
It includes taxonomy of human\hyp{}compatible spatial concepts which are linked to the sensed 
instances of these concepts drawn from lower layers. It is the conceptual layer which 
contains the information that kitchens commonly contain cereal boxes and have certain 
general appearance and allows the robot to infer that the cornflakes box in front of the 
robot makes it more likely that the current room is a kitchen. The conceptual layer is 
described in terms of a probabilistic ontology defining spatial concepts and linking 
those concepts to instances of spatial entities (see the example of the ontology in
\autoref{fig:dora-architecture}).

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.75\textwidth]{figures/chaingraph.pdf}
\caption{\label{fig:chain-graph}Example chain-graph produced by the \emph{conceptual layer}.}
\end{figure}

@@

\subsection{Factor Graphs}
Although the conceptual layer works with \emph{chain graphs}~\cite{lauritzen2002chain},
those can be converted into \emph{factor graphs}~\cite{kschischang2001factor}.
Factor graphs have been introduced in \autoref{sec:graphical-models} and are used
throughout this thesis as they provide a easier manipulation due to factorization.

Moreover, there exists efficient implementation of inference engines operating on factor
graph representations~\cite{Mooij_libDAI_10}.
Describing the distribution function in terms of graphs allows to use those engines to
efficently calculate marginals on any given subset of variables by exploiting conditional
independence between variables.

\subsection{Uncertain Sensing and Implicit Factors}
\label{sec:cues-from-low-level}
In a realistic scenario classification tasks are by their nature uncertain.
And although a trained classifier is able to decide which class is more likely, there is
interest in handling uncertainty produced by those to turn decisions more robust.

By modelling the classifiers as an implicit function $\phi_{classifier}(c, x)$ it becomes
possible to model the uncertainty of those given the presence of the sensed low-level
features $X$. The presence of those implicit models no longer allow to calculate the
normalization factor over the whole graph as there is no explicit model for either the
low\hyp{}level features $X$ (e.g.\ $X$ can be an image received from a visual sensor)
or for $\phi_{classifier}$ (e.g.\ the classifier can be an \gls{SVM}).

Nonetheless it is still possible to calculate any probability given that the features $X$
are observed. For that instead of modelling the variables $X$ and explicitly describing
the factor $\phi_{classifier}$ it is enough to replace it by an observed factor $\phi_{classifier}(c|x)$.

>>>>

% motivation
Humans associate concepts to areas that relate to the functionality of those. That
association is either useful by explaining where to find certain objects as well to
understand the types of activities to perform on it. In a

Those human-concepts are meant to be meaningful for humans and by incorporating conceptual
knowledge about them it is expected to improve the agent 
% semantic mapping definition
In this context semantic mapping is seen as the process of building a representation of
the environment which associates spatial entities with a set of defined spatial concepts (i.e.\ human concepts).

% importance of conceptual knowledge
% application of conceptual knowledge
It is expected that by being able to reason over those semantics an agent can better
understand human environments and increase its efficiency when performing human-tasks on them.
By handling this conceptual knowledge the agent becomes aware of

% challenges and need for representing uncertainty
The goal of a spatial representation is to provide the agent with a view about the environment
that allows it to reason beyond its sensory horizon.

 Such a representation is by its nature
imperfect, incomplete, inaccurate and invalid.

world that allows the agent to perform reasoning over it and efficiently perform tasks.
It should also give the agent ability to reason about parts 
Such a representation cannot be perfect since the methods and information are uncertain

% here a here a method is used that does semantic mappping and representes conceptual knowledge
% on a probabilistic fashion.

@@

This thesis motivation lies on developing methods for detecting knowledge gaps
on the semantic knowledge of a mobile agent. For that the semantic mapping system
implemented on \gls{Dora} proposed by \cite{pronobis2011phd} was used as a base.

\section{Architecture Overview}
Pronobis~\cite{pronobis2011phd} presents a system architecture working in indoor environments using
non\hyp{}omnidirectional laser and visual sensors for semantic mapping. The system integrates cues such as geometry,
object presence and appearances and is able to perform inference across any semantic properties,
for example for the purpose of place categorization.
The introduced system is built around a spatial knowledge representation~\cite{pronobis2010ias} and it has not only been tested on real-scenarios
and performance tested across several conditions~\cite{pronobis2007iros} but also shown
to be tailored to effectively solve tasks arising on mobile robotics~\cite{hanheide2011ijcai}.

Besides that the spatial knowledge representation has been drawn with an high-focus on probabilistic
and uncertain reasoning, human interaction and life-long learning capabilities.
For that it was considered an excellent base defining not only the concepts and
requirements of an high-level semantic representation but also the information flow
between all the layers involved on the agent.

Being able to perform probabilistic inference on all available information Such a representation is suitable to detect novel concepts.

The system is implemented on \Gls{Dora} and a brief overview of its architecture is
given in the following sections.

\section{Spatial Knowledge Representation}
\cite{pronobis2010ias} proposes a spatial knowledge representation sub-divided into four layers.
Each layer focus on different aspects of the world, abstractions levels of the spatial knowledge
and different spatial scales. Each uses different spatial entities and relate to the agent
goals in different ways:

In the lowest abstraction level the sensory layer represents the most immediate and short term
accurate representation of the world. Above, the place layer discretizes the continuous
space into. The categorical layer focus on knowledge of low\hyp{}level, long\hyp{}term
categorical models of the sensory information. And at the higher level the conceptual map
associates concepts to areas such as rooms (see \autoref{fig:spatial-knowledge}).

\begin{figure}[h]
\centering
\includegraphics[width=0.7\textwidth]{figures/dora-architecture.pdf}
\caption{\label{fig:spatial-knowledge}Layered structure of the spatial representation.
         The conceptual layer illustrates part of the ontology representing both instance
         and pre-defined world knowledge.}
\end{figure}

\subsubsection*{Sensory Layer}
The sensory layer keeps a detailed representation of the immediate surrounds of the agent.
That representation is based on directly sensed inputs as well on data-fusion for short
time\hyp{}intervals.

It is responsible for providing the agent with an exact position and to accurately position
low\hyp{}features and landmarks. The information is stored with associated uncertainty it
is highly susceptible to be replaced with newer versions. Also old or distant information
is forgotten.
By tracking the surrounds for short amount of times it provides an accurate representation
that can be used to locate and guide the agent on low\hyp{}level movements.

\subsubsection*{Place Layer}
The place layer is responsible for maintaining a bottom\hyp{}up discretization of the
continuous space. On it the world is represented as a collection of discrete spatial
entities, called places, together with connectivity information.

Each place is defined with the features from the sensory layer and with spatial relations
to other places. Connectivity here, stands for ability to travel directly between the
places. Additionally places can be created for areas not yet explored allowing to introduce
virtual place holders to places that would eventually be uncover if the agent moved there.

The places are also considered stable on long\hyp{}term and can be used to help the agent
performing localization and planning longer distance motion planning. This higher level
representation although uncertain allows to model connectivity and relaxed spatial relations.

\subsubsection*{Categorical Layer}
The categorical layer holds the generic long-term, low-level representations of categorical models
of the agent's sensory information.
The knowledge represented by this layer is not instance specific. It is the knowledge
required to determine certain features indicate the presence of an entity of certain
category. E.g.\ the knowledge required to detect objects, landmarks or describe room appearances.
Other properties may as well be defined for example shape, color, edges and size properties
are all defined in terms of low\hyp{}level features.

This layer is responsible for representing knowledge on how to extract properties used
by the conceptual layer.
Although not mandatory, in many cases the categories represented by it will map to human concepts
and not to the internal concepts of the agent.
The type of properties handled by this layer may require incredible complex models and for
that they might be trained on a supervised fashion.

\subsubsection*{Conceptual Layer}
The conceptual layer provides an ontology that represents a taxonomy of the spatial concepts
and their relations with the properties represented on the categorical layer.
This associates semantic meaning to those properties that is useful for human-agent iteration
and provides relations that can be verbalized and explained with the ontology.

It provides also knowledge between the semantic concepts and their instances of those concepts.
Including definitions of spatial concepts related to space segmentation as well to semantic
categorization of those spatial entities.
E.g.\ rooms are commonly split by doors, rooms exist in a floor, a building has floors, milk is
likely to be found on kitchens, etc\dots

By providing information for segmentation, semantic mapping to concepts and relations between those
concepts, the knowledge represented on this layer allows the agent to explain and reason close to
the human concepts. The knowledge represented here is considered to be valid through very long
periods of time or even life\hyp{}long.

\section{System Structure}
The \Gls{Dora} system~\cite{hanheide2011ijcai} consists of several co-operating
sub-systems, all of which actively use or maintain the spatial knowledge representation
(see \autoref{fig:dora-architecture}).

\subsection{Features}
\label{sec:dora-features}
The \emph{categorical layer} requires classifiers with the ability to categorize

\section{Conceptual Map}
\label{sec:conceptual-map}
As \gls{Dora} moves through the environment it builds a \emph{conceptual map}: a structural and
probabilistic representation of the space instantiated as a \emph{graphical model}.
It includes taxonomy of human\hyp{}compatible spatial concepts which are linked to the sensed 
instances of these concepts drawn from lower layers. It is the conceptual layer which 
contains the information that kitchens commonly contain cereal boxes and have certain 
general appearance and allows the robot to infer that the cornflakes box in front of the 
robot makes it more likely that the current room is a kitchen. The conceptual layer is 
described in terms of a probabilistic ontology defining spatial concepts and linking 
those concepts to instances of spatial entities (see the example of the ontology in
\autoref{fig:dora-architecture}).

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.75\textwidth]{figures/chaingraph.pdf}
\caption{\label{fig:chain-graph}Example chain-graph instantiated from the \emph{conceptual layer}.}
\end{figure}

@@

\subsection{Factor Graphs}
Although the conceptual layer works with \emph{chain graphs}~\cite{lauritzen2002chain},
those can be converted into \emph{factor graphs}~\cite{kschischang2001factor}.
Factor graphs have been introduced in \autoref{sec:graphical-models} and are used
throughout this thesis as they provide an easier manipulation due to factorization.

Moreover, there exist efficient implementations of inference engines operating on factor
graph representations~\cite{Mooij_libDAI_10}.
Describing the distribution function in terms of graphs permits the use of those engines to
efficently calculate marginals on any given subset of variables by exploiting conditional
independence between variables.

\subsection{Uncertain Sensing and Implicit Factors}
\label{sec:cues-from-low-level}
In a realistic scenario, classification tasks are by their nature uncertain.
And although a trained classifier is able to decide which class is more likely, there is
interest in handling uncertainty produced by those to make decisions more robust.

By modelling the classifiers as an implicit function $\phi_{classifier}(c, x)$ it becomes
possible to model the uncertainty of the decision given the presence of the sensed low-level
features $x$. The presence of those implicit models no longer allow calculation of the
normalization factor over the whole graph as there is no explicit model for either the
low\hyp{}level features $X$ (e.g.\ $X$ can be an image received from a visual sensor)
or for $\phi_{classifier}$ (e.g.\ the classifier can be an \gls{SVM}).

Nonetheless it is still possible to calculate any probability given that the features $X$
are observed. For that instead of modelling the variables $X$ and explicitly describing
the factor $\phi_{classifier}$ it is enough to replace it by an observed factor
$\phi_{classifier}(c|x)$ as shown on \autoref{fig:implicit-factor}.
